---
title: "R-Ladies East Lansing"
subtitle: dplyr key
output: html_document
---


## Load libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(nycflights13)
```

### Import data
```{r}
flights <- flights
flights

weather <- weather
weather
```

We can take a look at our data using the `head()` command.
```{r}
head(flights) 
```

### Using `filter()` to isolate rows based off of values


Find all flights departing May 9th
```{r}
may_9 <- flights %>% 
  filter(month == 5 & day == 9)

may_9
```

Find all flights in January and February
```{r}
jan_feb <- flights %>% 
  filter(month %in% c("1", "2"))

jan_feb
```


Find all flights to LAX and SFO.
```{r}
LAX_SFO <- flights %>% 
  filter(dest %in% c("SFO", "LAX"))

LAX_SFO
```


Find all flights delayed by > 60min.

```{r}
flights %>% 
  filter(arr_delay >60)
```

Find allflights that departed between 12 am and 6 am.

```{r}
early_flights <- flights %>% 
  filter(sched_dep_time >= 2400 |
         sched_dep_time <= 600 )

early_flights
```




### Using `arrange()` to sort rows


We can use `arrange()` to sort alphabetically.
```{r}
flights %>% 
  arrange(dest)
```

We can use `arrange()` to sort numeric values.
```{r}
flights %>% 
  arrange(air_time)
```

We can use `desc()` to sort in descending order.
```{r}
flights %>% 
  arrange(desc(air_time))
```



### Using `mutate()` to add a new column from existing data.
This function is great for performing calculations.



Create a new column named `speed` to calculate miles per hour (using `times` and `miles` columns.

```{r}
speed <- flights %>% 
  mutate(speed = distance/(air_time/60)) %>% 
  arrange(desc(speed))

speed
```


What was the longest flight delay in JFK in November?

```{r}
flights %>% 
  filter(month == 11 & origin == "JFK") %>% 
  arrange(desc(dep_delay))
```


Which flights departed from LGA arrived to DTW early?

```{r}
flights %>% 
  filter(origin == "LGA" & dest == "DTW") %>% 
  arrange(arr_delay)
```



Using `group_by()` and `summarize` to aggregate data.

We can combine multiple rows of data into a single value using `group-by` and `summarize`. 


Which airport had the most flights in December?

```{r}
flights %>% 
  filter(month == 12) %>% 
  group_by(origin) %>% 
  summarise(N_flights = length(flight))
```


Which NYC airport has the most airlines?

```{r}
flights %>% 
  group_by(origin) %>% 
  summarise(N_airlines = n_distinct(carrier))
```


How many United Airlines flights depart from JFK to ORD?
```{r}
flights %>% 
  filter(origin == "EWR" & carrier == "UA" & dest == "ORD") %>%
  summarise(N_flights = length(flight))
```


### Joining data
We can use `inner_join()` to combine data based off of similar values.


```{r}
weather_flights <- flights %>% 
  inner_join(weather)

weather_flights
```



### Advanced dplyr

We can combined what we've learned to answer some interesting questions.


Is there a relationship between flight arrival delay and overall trip distance? Or put more simply, Do longer distance flights have a shorter or longer arrival delay?

To answer this question we can aggregate the data using `group-by()` to group by destination and then use `summarize()` to aggregate the number of flights, mean distance and arrival delay. We next can use `filter()` to get the top 10 flights and since Hawaii is far from NYC we can omit it as well.

We can then use ggplot to plot the data.  We can make the `x axis` distance, the `y axis` arrival delay and the total number of flights the `size`.  We can also use `geom_smooth()` to fit a linear model to the data.

```{r}
# Code adapted from Hadley Wickham

delays <- flights %>%
  group_by(dest) %>% 
  summarise( count = n(),
             dist = mean(distance, na.rm = TRUE),
             delay = mean (arr_delay, na.rm = TRUE)) %>% 
  filter(count > 20 & dest != "HNL")



delays %>% 
  ggplot(aes(x= dist, y= delay))+
  geom_point(aes(size = count, alpha = 1/3))+
  geom_smooth(se = FALSE)
```

We can see that it appears that there is a slight increase in arrival delays for shorter flights but the delay decreases as you increase the trip distance.  